---
title: "Keystrokes_clean code"
author: "Andrea Camasta"
date: "2026-05-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Modelling Keystrokes

## Loading the dataset and the required packages

```{r}
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(see)
library(patchwork)
library(lme4)

data <- read_csv("data/data.csv")
```


## Factorization and Ordering of categorical variables

```{r}
data$`How good was the quality of MT?` <- factor(
  data$`How good was the quality of MT?`, 
  levels = c("Insufficient", "Fair", "Good", "Very good"),
  ordered = TRUE
)

data$`How difficult to translate was the source text?` <- factor(
  data$`How difficult to translate was the source text?`, 
  levels = c("Very difficult", "Somewhat difficult", "Neutral",
             "Somewhat easy", "Very easy"),
  ordered = TRUE
)

data$`How useful were the error annotations?` <- factor(
  data$`How useful were the error annotations?`, 
  levels = c("nan", "Extremely not useful", "Somewhat not useful",
             "Neutral", "Somewhat useful", "Extremely useful"),
  ordered = TRUE
)

data$`How useful were the error annotations?` <- factor(
  data$`How useful were the error annotations?`, 
  levels = c("nan", "Extremely not useful", "Somewhat not useful",
             "Neutral", "Somewhat useful", "Extremely useful"),
  ordered = TRUE
)

data$`How accurate were the error annotations?` <- factor(
  data$`How accurate were the error annotations?`, 
  levels = c("nan", "Somewhat inaccurate", "Not accurate nor inaccurate",
             "Somewhat accurate"),
  ordered = TRUE
)

data$`How useful were the translation suggestions?` <- factor(
  data$`How useful were the translation suggestions?`, 
  levels = c("nan", "Neutral", "Somewhat useful",
             "Extremely useful"),
  ordered = TRUE
)

data$`How accurate were the translation suggestions?` <- factor(
  data$`How accurate were the translation suggestions?`, 
  levels = c("nan", "Somewhat inaccurate", "Not accurate nor inaccurate",
             "Somewhat accurate", "Very accurate"),
  ordered = TRUE
)


data$PET <- as.factor(data$PET)
data$condition <- as.factor(data$condition)
data$domain <- as.factor(data$domain)
data$text_name <- as.factor(data$text_name)
data$Sent_id <- as.factor(data$Sent_id)

data$MT_Quality <- as.numeric(data$`How good was the quality of MT?`)
data$Difficulty <- as.numeric(data$`How difficult to translate was the source text?`)
data$condition <- factor(data$condition)
```


## Y variable Distribution

# raw data
ggplot(data, aes(x = keystrokes)) +
  geom_histogram(aes(y = ..density..),
                 bins = 30,
                 fill = "lightblue",
                 color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Keystrokes",
       x = "Keystrokes",
       y = "Density")



## Tweedie GLMM 

# The Tweedie distribution is a member of the Exponential Dispersion Family. Here, it is specifically a 
# Compound Poisson-Gamma distribution

# It acts as a single mathematical framework that handles two processes at once:
# 1) The Discrete Part (Poisson) models the probability of an event occurring (e.g., whether the 
#  translator chooses to edit the sentence or leave it at 0 keystrokes);
# The Continuous Part (Gamma) models the magnitude of the event (e.g., if they do edit, how many keystrokes are required)



library(glmmTMB)
full_model <- glmmTMB(keystrokes ~ condition * domain + num_characters + num_minor + num_major + MT_Quality + Difficulty + 
                   (1 | PET) + (1 | text_name) + (1 | Sent_id), 
                 family = tweedie(), data = data, REML = FALSE)
summary(full_model)

full_model$sdr$pdHess


library(DHARMa)
res <- simulateResiduals(full_model)
plot(res)

# A more simplified model
refined_model <- glmmTMB(keystrokes ~ condition * domain + 
                             num_characters + num_minor + num_major + 
                             MT_Quality + Difficulty + 
                             (1 | PET) + (1 | text_name), 
                           family = tweedie(), 
                           data = data)

summary(refined_model)


library(glmmTMB)

# 1. Scale variables to get numerical stability
final_recovery_model <- glmmTMB(
  keystrokes ~ condition * domain + 
    scale(num_characters) + 
    scale(num_minor) + 
    scale(num_major) + 
    scale(MT_Quality) + 
    scale(Difficulty) + 
    (1 | PET) +
    (1 | text_name / Sent_id),       
  family = tweedie(), 
  data = data
)

# 2. Check stability
library(DHARMa)
final_recovery_model$sdr$pdHess
res <- simulateResiduals(final_recovery_model)
plot(res)

summary(final_recovery_model)



#### visualising random intercepts ####
library(glmmTMB)
library(sjPlot)

plot_model(final_recovery_model, type = "re")




#### Simplifying the model to make residuals homoschedastic
# Creating the simplified version: no interaction term
reduced_model <- update(final_recovery_model, . ~ . - condition:domain)

# Compare them
anova(final_recovery_model, reduced_model)
# no significant differences in predictive power and AIC and BIC are lower, we keep the reduced

# Check Variance Inflation Factors
performance::check_collinearity(final_recovery_model) # since the full model has interaction terms. VIFs might be affected
performance::check_collinearity(reduced_model) # here no highly correlated variables


# 2. Check stability
library(DHARMa)
final_recovery_model$sdr$pdHess
res <- simulateResiduals(reduced_model)
plot(res)


# Let's try to simplify some more
summary(reduced_model)


# no Difficulty term
reduced_model_noDiff <- update(reduced_model, . ~ . - scale(Difficulty))

res <- simulateResiduals(reduced_model_noDiff)
plot(res)+
  labs(title = "Diagnostics plots - Keystrokes")
# now it works

summary(reduced_model_noDiff)


# Compare them
anova(reduced_model, reduced_model_noDiff)
# clear indication to stick with the second reduced model (no difficulty)

summary(reduced_model_noDiff)

#### Thus, the final model is reduced_model_noDiff

# Plotting the regressors' effects on keystrokes

library(ggplot2)
library(dplyr)

# extract fixed effects table
coefs <- summary(reduced_model_noDiff)$coefficients$cond
df <- as.data.frame(coefs)

# add term names
df$term <- rownames(df)

# rename columns for convenience
df <- df %>%
  rename(
    estimate = Estimate,
    std.error = `Std. Error`,
    p.value = `Pr(>|z|)`
  )

# get confidence intervals
ci <- confint(reduced_model_noDiff, method = "Wald")

# keep only fixed effects
ci <- as.data.frame(ci)
ci$term <- rownames(ci)

df <- df %>%
  left_join(ci, by = "term") %>%
  rename(conf.low = `2.5 %`, conf.high = `97.5 %`)

# remove intercept
df <- df %>%
  filter(term != "(Intercept)")

# relabel terms
df <- df %>%
  mutate(
    term = recode(term,
                  "condition2" = "H-QE",
                  "condition3" = "H-APE",
                  "condition4" = "S-APE",
                  "domainnews" = "Domain: News",
                  "scale(num_characters)" = "Text length",
                  "scale(num_minor)" = "Minor errors",
                  "scale(num_major)" = "Major errors",
                  "scale(MT_Quality)" = "MT quality"
    ),
    significance = ifelse(p.value < 0.05, "Significant", "Not significant")
  )

library(ggplot2)
library(dplyr)
library(forcats)

# reorder by significance first, then by effect size
df <- df %>%
  mutate(significant_flag = p.value < 0.05) %>%
  arrange(desc(significant_flag), desc(abs(estimate))) %>%
  mutate(term = factor(term, levels = rev(term)))

# plot

p <- ggplot(df, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.6, color = "gray50") +
  
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.2, linewidth = 0.8, color = "gray40") +

  geom_point(aes(shape = as.factor(!significant_flag)),
             size = 3, stroke = 1.1, color = "black") +
  
  scale_shape_manual(
    name = NULL,              
    values = c(16, 1),        
    labels = c("Significant", "Not Significant") 
  ) +
  
  labs(
    title = "Effects on Keystrokes",
    subtitle = "Estimates with 95% confidence intervals (log scale)",
    x = "Effect size",
    y = NULL
  ) +
  
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray30"),
    axis.text.y = element_text(size = 11),
    axis.text.x = element_text(size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray85"),
    plot.margin = margin(10, 15, 10, 10),
    legend.position = "right"
  )

p



# save 
#ggsave(
#  filename = "keystrokes_effects_plot.png",
#  plot = p,
#  width = 7,
#  height = 4.5,
# dpi = 300
#)


#### random intercepts plots 
library(ggplot2)
library(glmmTMB)
library(sjPlot)

sjPlot::plot_model(reduced_model_noDiff, type = "re", grid = FALSE)[[2]]+
  theme_minimal()+
  labs(x = "Text", y= "Random Intercept", title = "Random Effects: Texts")




#### Comparing final model Vs other models

# Definin main formula
main_formula <- keystrokes ~ condition + domain + scale(num_characters) + 
  scale(num_minor) + scale(num_major) + scale(MT_Quality) + 
  (1 | PET) + (1 | text_name / Sent_id)



# Current Tweedie Model
mod_tweedie <- glmmTMB(main_formula, 
                       family = tweedie(link = "log"), 
                       data = data)


# Zero-Inflated Negative Binomial
# (Assumes some zeros come from a separate always-zero process)
mod_zinb <- glmmTMB(main_formula, 
                    ziformula = ~1, 
                    family = nbinom2, 
                    data = data)

# 3. Hurdle Model (Truncated Negative Binomial)
# (Assumes a process: 0 vs >0, then counts for the rest)
mod_hurdle <- glmmTMB(main_formula, 
                      ziformula = ~., 
                      family = truncated_nbinom2, 
                      data = data)

library(performance)

# Compare them side-by-side
comparison <- compare_performance(mod_tweedie, mod_zinb, mod_hurdle, metrics = "common")
print(comparison)


# The 'rho' (power parameter)
glmmTMB::family_params(mod_tweedie)

# Even though the Hurdle model has a slightly lower AIC, 
# We would argue for the Tweedie model for three reasons:
# 1) The BIC weight for the Tweedie model is >.999. This is a massive statistical signal that the Tweedie model
# is the most parsimonious. 
# 2) The Tweedie model has the lowest RMSE (31.539). This means its actual predictions are closer to actual data points
# than the other two models.
# 3) In the Tweedie model, the Itra Class Correlation (ICC) is 0.548, while in the ZI model, it's 0.956, which is
# high and might suggests that the model is struggling to separate the random effects from the zero-inflation logic.

# For the Marginal R^2 (0.191): Fixed effects (condition, length, quality, etc.) explain about 19.1% of the variance 
# in keystrokes. 
# For Conditional R^2 (0.634): Adding the random part we are explaining nearly 64% of the variance.
